---
title: "Personalized Medicine: Exploration"
output: 
  html_document:
    keep_md: true
---
In this notebook, I will just explore the data and try to get some insights on how the target variable `Class` is related to the features (extracted or engineered from the data). 

```{r, echo=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(tidytext)
library(tokenizers)
library(grid)
library(gridExtra)
library(ggplot2)
```
We first load the variants data, which is simple:
```{r}
train_variants <- read_csv("./data/training_variants", progress = FALSE)
test_variants <- read_csv("./data/test_variants", progress = FALSE)
```
Loading the text data is a little more involved. A simple call to `read_delim` was unsuccesful. So we read the data as a simple text and then split into a data frame as follows:
```{r}
# train
train_text <- read_lines("./data/training_text", skip=1)
train_text_df <- str_split_fixed(train_text, pattern = "\\|\\|", n = 2) %>% tbl_df()
colnames(train_text_df) <- c("ID", "Text")
train_text_df <- train_text_df %>% mutate(ID = as.integer(ID))
# test
test_text <- read_lines("./data/test_text", skip=1)
test_text_df <- str_split_fixed(test_text, pattern = "\\|\\|", n = 2) %>% tbl_df()
colnames(test_text_df) <- c("ID", "Text")
test_text_df <- test_text_df %>% mutate(ID = as.integer(ID))
```
```{r, echo=FALSE}
rm(train_text,test_text); gc()
```

### Variants data

Let's look at how `Class` is distributed in training data (this is the target):
```{r}
train_variants %>%
  group_by(Class) %>%
  count(Class,sort=TRUE)
```
Clearly, there is an inbalance in the class labels, which we need to be aware of. 

#### Gene
Let's look at how `Gene` feature is distributed. The top 25 most common genes for training/test data are:
```{r}
df0 <- train_variants %>% 
  group_by(Gene) %>%
  count(Gene,sort=TRUE) %>% 
  slice(1:25)

df00 <- test_variants %>%
  group_by(Gene) %>%
  count(Gene,sort=TRUE) %>%
  slice(1:25)
```

```{r, fig.width=10, fig.height=5}
gg0 <- ggplot(df0, aes(reorder(Gene, -n, FUN=min), n)) + 
  geom_bar(stat="identity", fill = "salmon") + 
  coord_flip() + 
  labs(x = "Gene", y = "Count") + ggtitle('Training Data')

gg00 <- ggplot(df00, aes(reorder(Gene, -n, FUN=min), n)) + 
  geom_bar(stat="identity", fill = "salmon") + 
  coord_flip() + 
  labs(x = "Gene", y = "Count") + ggtitle('Test Data')

grid.arrange(gg0, gg00, ncol=2)

```
Clearly, there is rather a different distribution of top Genes in training and test data.
```{r}
intersect(df0$Gene, df00$Gene)
```
This is probably due to the artificial data generated in the test set to avoid hand labelling.

How is `Gene` related to Class (outcome)
```{r}
top_genes <- df0 %>% filter(n > 40)
df1 <- train_variants %>%
  filter(Gene %in% top_genes$Gene) %>%
  group_by(Gene,Class) %>%
  summarise(count = n()) 

gg1 <- ggplot(df1, aes(Gene,count)) + 
  geom_bar(stat = 'identity', fill='salmon') + 
  theme(axis.text.x  = element_text(angle=90)) +
  facet_wrap(~Class)
gg1
```

#### Variation
How does `Variation` is distributed? 
```{r}
df1 <- train_variants %>% 
  group_by(Variation) %>%
  count(Variation, sort=TRUE) %>%
  mutate(is.train = TRUE) 


df11 <- test_variants %>% 
  group_by(Variation) %>%
  count(Variation, sort=TRUE) %>%
  mutate(is.train = FALSE)


df111 <- full_join(df1, df11, by = c("Variation", "n", "is.train"))
df111[is.na(df111)] <- 0
df111 %>% top_n(8, n) 
```
There are so many unique values to `Variation`, and most of them are observed once.

How is `Variation` is related to clas?
```{r}
top_variations <- df111 %>% filter(n > 3)
df2 <- train_variants %>%
  filter(Variation %in% top_variations$Variation) %>%
  group_by(Variation,Class) %>%
  summarise(count = n()) 

gg2 <- ggplot(df2, aes(Variation,count)) + 
  geom_bar(stat = 'identity', fill='salmon') + 
  theme(axis.text.x  = element_text(angle=90)) +
  facet_wrap(~Class)
gg2
```
Truncating mutations seem to have a large correlation with Class = 1. None of the top variations has any knowledge of Class 3,4,9.

### Ratio features
Let's engineer ratio features (`n_Variatiob_Class`/`n_Variation`) using 5 folds. See documentation for details. 
```{r}
source('create_ratio_features.R')
train_new <- create_ratios_folds(train_variants, 5)
names(train_new) <- c("ID", "Gene", "Variation", "Class", paste0("R",1:9))
```
Let's look at this data
```{r}
train_new
```
How does the new ratio features (`R1`,.., `R9`) correlate with Class?
```{r}
gg3 <- ggplot(train_new %>% gather(key = "Ratio", value = "value", -c(ID,Gene,Variation,Class)) %>% 
                mutate(Ratio = as.factor(Ratio)) %>%
                mutate(Class = as.factor(Class)) %>%
                filter(Ratio %in% c("R1","R4", "R9")), 
              aes(value)) + 
  geom_histogram(aes(fill = Ratio), binwidth = 0.1) + 
  facet_wrap(~Class) + 
  ggtitle("R1,R4,R9 Counts for Classes")
gg3
```
Clearly, these ratio features correlate better with the Class they represent.

### Text data
```{r,echo=FALSE}
rm(train_new,df1,df11,df111,df0,df00,df2); gc()
```
Now let's tokenize train
```{r}
train_txt <- train_text_df %>% unnest_tokens(word, Text)
```
Most common words
```{r}
train_txt %>% count(word, sort=TRUE) %>% slice(1:10)
```
These are stop words that we can remove. 

```{r}
library(SnowballC)
data("stop_words")
# Extra words that can be removed which we add by hand
my_stopwords <- data_frame(word = c(as.character(1:100),
                                    "fig", "figure", "et", "al", "table",
                                    "data", "analysis", "analyze", "study",
                                    "method", "result", "conclusion", "author",
                                    "find", "found", "show", "perform",
                                    "demonstrate", "evaluate", "discuss"))
train_txt <- train_txt %>% mutate(word = wordStem(word))
train_txt <- train_txt %>%
  anti_join(stop_words, by = 'word') %>%
  anti_join(my_stopwords, by = 'word') %>%
  filter(str_detect(word, "[:alpha:]")) # No numbers
```
```{r}
train_txt %>% count(word, sort=TRUE) %>% slice(1:20)
```
Let's visualize
```{r}
gg4 <- ggplot(train_txt %>% count(word, sort=TRUE) %>% slice(1:25), aes(word,n)) + 
  geom_bar(stat="identity", fill='salmon') + 
  coord_flip()
gg4
```
Use term frequency-inverse document frequency (tf-idf) function. 
First, treat "Class" as "Document" of td-idf: 
```{r}
foo <- train_variants %>% select(ID, Class)
txt_class <- full_join(train_txt, foo, by = "ID")

frequency <-txt_class %>%
  count(Class, word)

tf_idf <- frequency %>%
  bind_tf_idf(word, Class, n)
```
```{r}
tf_idf %>% filter(tf_idf > 1e-3)
```
This gives us a subset of words that are correlated with class labels.

In the standard td-idf approach, the "Document" should be `ID`. Let's try that as well:
```{r}
foo <- train_variants %>% select(ID, Class)
txt_class <- full_join(train_txt, foo, by = "ID")

frequency <-txt_class %>%
  count(ID, word, sort=TRUE) %>%
  slice(1:5000) # Keep only the first 5000 most frequent words

tf_idf <- frequency %>%
  bind_tf_idf(word, ID, n)
```
```{r}
tf_idf %>% filter(tf_idf > 1e-3)
```
I'm not sure why we get the above warning, but the function seem to work. Looks like treating `Class` as document seems like a better idea.
